---
title: "04_DESeq2_macs"
output: html_document
date: "2023-11-11"
---

```{r}
library(tidyverse)
library(cowplot)
library(edgeR)
library(reshape2)
library(SingleCellExperiment)
library(pheatmap)
library(apeglm)
library(png)
library(DESeq2)
library(RColorBrewer)
library(data.table)
library(FactoMineR)
library(factoextra)  
library(tidyverse)
library(pheatmap)
library(DESeq2)
library(RColorBrewer)
library(viridis)  
library(ggplot2)
library(grid)
library(ggrepel)
library(ggrepel)
library(Cairo)
```

#PCA
```{r}
# Transform counts for data visualization
rld <- vst(DESeq, blind=TRUE)

# Plot PCA
DESeq2::plotPCA(rld, ntop = 500, intgroup = c("group.id","sample.id"))
```

```{r}
#plot the PCA
pcaPlot <- plotPCA(rld, ntop = 500, intgroup = c("group.id","sample.id"), returnData = TRUE)
# Convert to ggplot object

# Set your desired figure size and font size
fig_width <- 6
fig_height <- 4
font_size <- 3 # Adjust as needed
point_size <- 3 # Adjust as needed

# Define colors for group.id (adjust the colors as needed)
group_colors <- c("neg" = "blue", "pos" = "red") # Replace with actual group names and desired colors

# Generate the plot with ggrepel for non-overlapping text labels
ggplot(pcaPlot, aes(x = PC1, y = PC2, color = group.id, label = sample.id)) +
    geom_point(size = point_size) +
    geom_text_repel(size = font_size, 
                    box.padding = 0.5, 
                    point.padding = 0.3,
                    segment.size = 0.2) +
    scale_color_manual(values = group_colors) +
    theme(legend.position="none", 
          text = element_text(size = font_size),
          panel.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    theme_minimal() +
    theme(axis.text.x = element_text(size = font_size),
          axis.text.y = element_text(size = font_size),
          axis.title = element_text(size = font_size))



ggsave(filename = "Teff-pca_by_group_id.pdf", width = 12, height = 9, dpi = 300, path = './outs/pseudo_bulk/pos_neg/neut/')
```

#Hierarchical clustering
```{r}
# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)

# Plot heatmap
h <- pheatmap(rld_cor, annotation = cluster_metadata[, c("group.id", "sample.id"), drop=F])
h
cairo_pdf(file = './pseudobulk_3.0/compare_positive_across_clusters_find_tissue_specific/outs/clustering.pdf', width = 8, height = 5)
print(h)
dev.off()
```

#Running DESeq2
```{r}
# Run DESeq2 differential expression analysis
DESeq <- DESeq(DESeq)
# Plot dispersion estimates
#We can check the fit of the DESeq2 model to our data by looking at the plot of dispersion estimates.
plotDispEsts(DESeq)
#ggsave(filename = "DEGseq_plot.pdf", width = 12, height = 9, dpi = 300, path = '../four_bone_combined_analysis/20221021_old_vs_young_macs_outs//bulk_DEGs_outs/')
```

#Results
```{r}
#Let’s compare the stimulated group relative to the control:
# Output results of Wald test for contrast for pos vs neg
#method 1 design the exp and ctl groups -----------------------------------------------------------------------------------------
unique(levels(cluster_metadata$group.id))

levels(cluster_metadata$group.id)[1]
levels(cluster_metadata$group.id)[2]
levels(cluster_metadata$group.id)[3]
levels(cluster_metadata$group.id)[4]

levels(cluster_metadata$group.id)[5]
levels(cluster_metadata$group.id)[6]
levels(cluster_metadata$group.id)[7]
levels(cluster_metadata$group.id)[8]
#the following contrast matrix can only compare two groups
contrast <- c("group.id", 
              levels(cluster_metadata$group.id)[8],
              #levels(cluster_metadata$group.id)[1],
              levels(cluster_metadata$group.id)[7]
              )

res <- results(Neuts, 
               contrast = contrast,
               alpha = 0.05)

res
#First let’s generate the results table for all of our results:
# Turn the results object into a tibble for use with tidyverse functions
res_tbl <- res %>%
        data.frame() %>%
        rownames_to_column(var="gene") %>%
        as_tibble()

# Check results output
res_tbl

# Write all results to file
write.csv(res_tbl,
          paste0("results/", clusters[1], "_", levels(cluster_metadata$sample)[2], "_vs_", levels(cluster_metadata$sample)[1], "_all_genes.csv"),
          quote = FALSE, 
          row.names = FALSE)


#method 2, design exp and ctl group by default comparisons----------------------------------------------------------------------------------------------
# Check the coefficients for the comparison
resultsNames(Neuts)
#[1] "Intercept"                 "group.id_DSS_vs_ctrl"      "group.id_DSS.B109_vs_ctrl"

# Generate results object
res <- results(Neuts, 
               name = "group.id_pos_vs_neg", 
               alpha = 0.05)

# Shrink the log2 fold changes to be more appropriate using the apeglm method - should cite [paper]() when using this method
res <- lfcShrink(Neuts, 
                 coef = "group.id_DSS_vs_ctrl",
                 res=res,
                 type = "apeglm")
```

#Table of results for all genes
```{r}
# Turn the results object into a tibble for use with tidyverse functions
res_tbl <- res %>%
        data.frame() %>%
        rownames_to_column(var="gene") %>%
        as_tibble()


# Check results output
res_tbl
res_tbl <- na.omit(res_tbl)
# Write all results to file
write.csv(res_tbl,
          quote = FALSE, 
          row.names = FALSE,
          './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c7/res_tbl.csv')

saveRDS(res_tbl, './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c7/res_tbl.rds')
```

Table of results for significant genes
```{r}
# Set thresholds
pvalue_cutoff <- 0.05

# Subset the significant results
sig_res <- dplyr::filter(res_tbl, pvalue < pvalue_cutoff) %>%
        dplyr::arrange(pvalue)

# Check significant genes output
sig_res

write.csv(sig_res,
          quote = FALSE, 
          row.names = FALSE,
          './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c7/sig_res.csv')

saveRDS(sig_res, './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c7/sig_res.rds')

```

```{r, normolize}
normalized_counts <- counts(Neuts, normalized = TRUE)
```

#Scatterplot of normalized expression of top 20 most significant genes
##Now that we have identified the significant genes, we can plot a scatterplot of the top 20 significant genes. This plot is a good check to make sure that we are interpreting our fold change values correctly, as well.
```{r}
## ggplot of top genes
##fitering parameters
log2FC_cutoff = log2(1.5)
pvalue_cutoff = 0.05

##select the differential gene list 
need_DEG <- sig_res[,c(1,3,6)] #select log2FoldChange, p value information
colnames(need_DEG) <- c('gene','log2FoldChange','p_value') 

need_DEG$significance  <- as.factor(ifelse(need_DEG$p_value < pvalue_cutoff & abs(need_DEG$log2FoldChange) > log2FC_cutoff, ifelse(need_DEG$log2FoldChange > log2FC_cutoff ,'UP','DOWN'),'NOT'))

title <- paste0(' Up :  ',nrow(need_DEG[need_DEG$significance =='UP',]) ,
                     '\n Down : ',nrow(need_DEG[need_DEG$significance =='DOWN',]),
                     '\n FoldChange >= ',round(2^log2FC_cutoff,3))

need_DEG$delabel <- NA
need_DEG$delabel[need_DEG$significance != "NOT"] <- row.names(need_DEG)[need_DEG$significance != "NOT"]

g <- ggplot(data=need_DEG, 
            aes(x=log2FoldChange, y=-log10(p_value), 
                color=significance, label=delabel)) +
  #dots and background
  geom_point(alpha=0.4, size=1) +
  theme_classic()+ #grid
  xlab("log2 ( FoldChange )") + 
  ylab("-log10 ( P.value )") +
  ggtitle( title ) +
  scale_colour_manual(values = c('blue','grey','red'))+ 
  scale_fill_brewer(palette="Set1")+
  geom_vline(xintercept = c(-log2FC_cutoff,log2FC_cutoff),lty=4,col="grey",lwd=0.5) +
  geom_hline(yintercept = -log10(pvalue_cutoff),lty=4,col="grey",lwd=0.5) +
  
  theme(plot.title = element_text(hjust = 0.5), 
        plot.margin=unit(c(2,2,2,2),'lines'), #top, right, bottom, left
        legend.title = element_blank(), 
        legend.position="right")
g

ggsave(g,filename = './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c7/vocanol_no_labeling.png',width =8.5,height =8.5, dpi= 600)

```

```{r, with dot labeled}
need_DEG$significance<- "NOT"

need_DEG$significance[need_DEG$log2FoldChange > 0.58 & need_DEG$p_value < 0.05] <- "UP"

need_DEG$significance[need_DEG$log2FoldChange < -0.58 & need_DEG$p_value < 0.05] <- "DOWN"


need_DEG %>% remove_rownames %>% column_to_rownames(var="gene")

need_DEG$delabel <- NA
need_DEG$delabel[need_DEG$significance != "NOT"] <- need_DEG$gene[need_DEG$significance != "NOT"]

v <- ggplot(data=need_DEG, aes(x=log2FoldChange, y=-log10(p_value), col=significance, label=delabel)) +
        geom_point() + 
        theme_minimal() +
        geom_text_repel() +
        scale_color_manual(values=c("blue", "grey", "red")) +
        geom_vline(xintercept=c(-0.6, 0.6),lty=2,col="grey",lwd=0.5) +
        #geom_hline(yintercept=-log10(0.05), lty=4,col="grey",lwd=0.8) +
    geom_point(alpha=0.4, size=1) +
  theme_classic()+ #grid
  xlab("log2 ( Fold Change )") + 
  ylab("-log10 ( P value )") +
  ggtitle( title ) +
  scale_colour_manual(values = c('blue','grey','red'))+ 
  geom_vline(xintercept = c(-log2FC_cutoff,log2FC_cutoff),lty=2,col="grey",lwd=0.5) +
  geom_hline(yintercept = -log10(pvalue_cutoff),lty=2,col="grey",lwd=0.5) +
  
  theme(plot.title = element_text(hjust = 0.5), 
        plot.margin=unit(c(2,2,2,2),'lines'), #top, right, bottom, left
        legend.title = element_blank(), 
        legend.position="right")

v

ggsave(v,filename = 'vocanol_with_labeling.png',width =10,height =10, dpi = 300, path = './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c5/')

```

```{r, Heatmap}
# Heatmap
## Extract normalized counts for significant genes only
sig_counts <- normalized_counts[rownames(normalized_counts) %in% sig_res$gene, ]

## Set a color-blind friendly palette
heat_colors <- rev(brewer.pal(10, "PuOr"))
breaks <- seq(-1.5, 1.5, length.out = length(heat_colors) + 1)
## Run pheatmap using the metadata data frame for the annotation
hp1 <- pheatmap(sig_counts, 
         color = heat_colors, 
         breaks = breaks,
         cluster_rows = TRUE, 
         show_rownames = FALSE,
         annotation = cluster_metadata[, c("sample.id", "group.id", "cell.id")], 
         border_color = NA, 
         fontsize = 7, 
         scale = "row", 
         fontsize_row = 10, 
         height = 20,
         angle_col = 45)

ggsave(hp1, filename = 'heatmap_sig_genes.png',width =10,height =18, dpi = 300, path = './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c5/')

```
```{r}
#make the plot color darker, adjust the color scale
# Function to make colors darker
sig_counts <- normalized_counts[rownames(normalized_counts) %in% sig_res$gene, ]

## Set a color-blind friendly palette
heat_colors <- rev(brewer.pal(7, "PuOr"))

# Define the breaks for the colors, ensuring to cover the range -4 to 4
breaks <- seq(-1.5, 1.5, length.out = length(heat_colors) + 1)

# Run pheatmap without the unsupported arguments
hp1 <- pheatmap(sig_counts, 
        color = heat_colors, 
        breaks = breaks,  # Use the defined breaks
        cluster_rows = TRUE, 
        show_rownames = FALSE,
        border_color = NA, 
        fontsize = 7, 
        scale = "row", 
        fontsize_row = 5)
hp1

cairo_pdf(file = './pseudobulk_2.0/outs/mye_grouped_by_tissue-biotin/Mφ.c3/heatmap_sig_genes.pdf', width = 6, height = 8)
print(hp1)
dev.off()
```








```{r}
library(pheatmap)
library(dplyr)
breaks <- seq(-1.5, 1.5, length.out = length(heat_colors) + 1)

# Assuming cluster_metadata is a data frame with "sample.id" matching the column names of sig_counts
# and sig_counts is a matrix with column names

# Calculate the order for columns based on "sample.id"
# This assumes that cluster_metadata is aligned with the columns of sig_counts
custom_order <- c(
  "Mφ_IC-bone-A-neg-rep1", "Mφ_IC-bone-A-neg-rep2", 
   "Mφ_IC-bone-Y-neg-rep1", "Mφ_IC-bone-Y-neg-rep2",
  "Mφ_IC-brain-A-neg-rep1", "Mφ_IC-brain-A-neg-rep2",
  "Mφ_IC-brain-Y-neg-rep1", "Mφ_IC-brain-Y-neg-rep2",
  "Mφ_IC-liver-A-neg-rep1", "Mφ_IC-liver-A-neg-rep2", 
   "Mφ_IC-liver-Y-neg-rep2", "Mφ_IC-liver-Y-neg-rep1",
  "Mφ_IC-lung-Y-neg-rep1", "Mφ_IC-lung-Y-neg-rep2",
   "Mφ_IC-lung-A-neg-rep2","Mφ_IC-lung-A-neg-rep1",
  
  
  "Mφ_IC-bone-A-pos-rep1", "Mφ_IC-bone-A-pos-rep2",
  "Mφ_IC-bone-Y-pos-rep1", "Mφ_IC-bone-Y-pos-rep2",
  "Mφ_IC-brain-A-pos-rep1", "Mφ_IC-brain-A-pos-rep2",
   "Mφ_IC-brain-Y-pos-rep1", "Mφ_IC-brain-Y-pos-rep2",
  "Mφ_IC-liver-A-pos-rep1",  "Mφ_IC-liver-Y-pos-rep1",
  "Mφ_IC-liver-Y-pos-rep2", 
  "Mφ_IC-lung-A-pos-rep1", "Mφ_IC-lung-A-pos-rep2",
  "Mφ_IC-lung-Y-pos-rep1", "Mφ_IC-lung-Y-pos-rep2"
)

# Match the custom order to the actual column names in sig_counts
column_order <- match(custom_order, colnames(sig_counts))

# Check if all sample IDs in custom_order are present in the column names
if (any(is.na(column_order))) {
  stop("One or more sample IDs in the custom order do not match the column names of sig_counts.")
}

# Reorder sig_counts columns according to the custom order
sig_counts_ordered <- sig_counts[, column_order, drop = FALSE]

# Assuming cluster_metadata has rows in the same order as the original sig_counts,
# reorder cluster_metadata to match the new column order.
cluster_metadata_ordered <- cluster_metadata[column_order, ]

# Run pheatmap with the ordered columns and the corresponding ordered metadata
hp2 <- pheatmap(sig_counts_ordered,
                color = heat_colors,
                breaks = breaks, # Make sure you have defined the 'breaks' variable earlier
                cluster_cols = F,
                cluster_rows = TRUE, 
                show_rownames = FALSE,
                annotation_col = cluster_metadata_ordered,
                border_color = NA, 
                fontsize = 7, 
                scale = "row", 
                fontsize_row = 5, 
                height = 15,
                angle_col = 90)

ggsave(hp2, filename = 'heatmap_sig_genes.IC.pdf',device = cairo_pdf, width =6,height =10, dpi = 300, path = './outs/pseudo_bulk/pos_neg/macs/')

```


